Changes in Patterns of Geometric Sign Groupings as a Model of Social Dynamics Across Aurignacian Phases

### 2025-06-02 paper/paper.qmd
map

In identifying the groups, we referenced both the seriation and network analyses. The results of the above tests allowed us to find two main groups of sites within each phase (@fig-seriation).

@@

Our first aim was to determine how the grouping dynamics of geometric signs changed across four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and distribution. The results from @fig-seriation show that each phase can be divided into two main cultural groups, with sign type diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian. @fig-neighbor-net shows that group inter-connectivity and intra-connectivity varies from phase to phase, with weak intra-group connections and no inter-group connections in the Transitional phase, to stronger intraconnections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian. The perMANOVA tests show that the changes between the the Proto, Early, and Evolved Aurignacian phases are statistically significant, that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian has the least. The

One interpretation of the inter-connectivity seen in @fig-neighbor-net is as a representation of non-utilitarian mobility. Non-utilitarian mobility establishes regional social networks which facilitate the spread of information and the survival of hunter-gatherers in uncertain environments [@Whallon_2006; @Fitzhugh_Phillips_Gjesfeld_2011; @Lovis_Donahue_2011; @Hitchcock_Ebert_2011; @Romano_Lozano_Fern2002]. The exchange and production of decorative items is associated with non-utilitarian mobility and has been recognized as a sign of social contact in other regions and eras [@Whallon_2006; @Kelly_2024; @Kinahan_2013; @Newlander_2012; @Iizuka_Ferguson_Izuho_2022]. Existing data has shown a range of variation in mobility among hunter-gatherer groups [@Binford_1990; @Kelly_1983; @Johnson_2014; @Padilla-Iglesias_Bischoff_2024; @Singh_Glowacki_2022]; some of the inter-connections in @fig-neighbor-net may represent shorter-term information exchange, while others may indicate longer-term residential movement.

One form of longer-term residential movement is the aggregation/dispersion pattern, which is defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for subsistence or social reasons. Aggregation sites, associated with high degrees of connectivity, are the places where those group fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg]. Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994]. The possibility that some sites functioned as aggregation sites may have influenced the geographic structure of sites (@fig-group-map), which changes from phase to phase.

>>>>

In identifying the groups, we referenced both the seriation and network analyses. The results of the above tests allowed us to find two main groups of sites within each phase (@fig-group-map).

@@

Our first aim was to determine how the grouping dynamics of geometric signs changed across four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and distribution. The results from @fig-seriation show that each phase can be divided into two main cultural groups, with sign type diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian. @fig-neighbor-net shows that group inter-connectivity and intra-connectivity varies from phase to phase, with weak intra-group connections and no inter-group connections in the Transitional phase, to stronger intraconnections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian. The perMANOVA tests show that the changes between the the Proto, Early, and Evolved Aurignacian phases are statistically significant, that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian has the least.

One interpretation of the connectivity seen in @fig-neighbor-net is as a representation of non-utilitarian mobility. Non-utilitarian mobility establishes regional social networks which facilitate the spread of information and the survival of hunter-gatherers in uncertain environments [@Whallon_2006; @Fitzhugh_Phillips_Gjesfeld_2011; @Lovis_Donahue_2011; @Hitchcock_Ebert_2011; @Romano_Lozano_Fern2002]. The exchange and production of decorative items is associated with non-utilitarian mobility and has been recognized as a sign of social contact in other regions and eras [@Whallon_2006; @Kelly_2024; @Kinahan_2013; @Newlander_2012; @Iizuka_Ferguson_Izuho_2022]. Previous studies have shown a range of variation in mobility among hunter-gatherer groups [@Binford_1990; @Kelly_1983; @Johnson_2014; @Padilla-Iglesias_Bischoff_2024; @Singh_Glowacki_2022]; some of the connections in @fig-neighbor-net may represent shorter-term information exchange, while others may indicate longer-term residential movement.

One form of longer-term residential movement is the aggregation/dispersion pattern, which is defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for subsistence or social reasons. Aggregation sites, associated with high degrees of connectivity, are the places where those group fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg]. Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994]. The possibility that some sites functioned as aggregation sites may have influenced the geographic structure of sites (@fig-group-map), which change from phase to phase.
